Optimal Neural Summarisation for Full-Field Implicit Inference by Density Estimation

### 2023-09-08 paper/paper.tex
text 08/09/2023

  \abstract
  % context heading (optional)
  % {} leave it empty if necessary  
   {In many cosmological applications, the likelihood function is either unknown or intractable, limiting the use of traditional inference methods. Likelihood-Free Inference (LFI) approaches offer a promising solution by using forward-simulated synthetic data to approximate the posterior (or likelihood) distribution instead of evaluating an explicit likelihood function. However, these methods suffer from the curse of dimensionality, necessitating the development of compression schemes to reduce high-dimensional data into lower-dimensional summary statistics.

   \justine{\textbf{I think the 2 main ideas for this part are: }

    - classical way to overcome this issue: we compress the data into 2pt statistics  and use gaussian likelihood → pb: we miss a lot of information}
    \denise{That's true. But adding it in the abstract would make the "Context" subsection (this division is mandatory for A$\&$A) too long. This is properly described in the introduction}

    \justine{ - considering the previous pb: thanks to Implicit Inference methods we can consider any kind of summary stat (because we can learn the exact posterior p(theta | summary stat)), so to have the best contours (meaning that we exatrct all gaussian as well as non-gaussian signal) we only have to focus on how we compress our data.}
    \denise{That's also true. But we don't use any summary statistics (except the power spectrum because it's easy), and I think that one of the points of the paper is to extract the information from the pixels, without using any summaries. So I personally would not talk about summary statistics }

     \justine{\textbf{Note:}

    - for me LFI = Implicit Inference, I think we should use only one
   }
   \denise{True again. But the definition "Likelihood-Free Inference" is more used in the community. Implicit Inference is something that we introduce in the motivation and introduction. I don't know exactly how much people are familiar with this concept. }
    \justine{ - 'Likelihood-Free Inference (LFI) approaches offer a promising solution by using forward-simulated synthetic data instead of an explicit likelihood function.' → Implicit inference approaches offer a promising solution to approximate posterior distribution by using simulations from an implicit likelihood (a simulator) instead of evaluating explicit likelihood function ..?

    - I'm probably wrong but not sure about the last sentence.
}
   }   
  % aims heading (mandatory)
    {We aim to derive the optimal neural-compression strategy for full field Likelihood free inference by density estimation. Moreover, by using the optimal compression strategy, we demonstrate that the posterior distribution is comparable to those derived from a Bayesian forward modeling approach.}  
  % methods heading (mandatory)
    { We present a comparative analysis of different loss functions employed during the training of the neural network. We  evaluate their performance by measuring their impact on the constraints on the $\Lambda$CDM parameters expected from LSST-Y10.
    For both forward-modeling the convergence field and generating observed mock data, we use the \href{https://github.com/DifferentiableUniverseInitiative/sbi_lens}{\url{SbiLens}} package. 
    \texttt{SbiLens} is a jax-based differentiable physical model specifically designed for Likelihood Free Inference and Bayesian inference algorithms that require access to likelihood derivatives.

    \justine{\textbf{I think the main ideas for this part are: }

    - first we describe sbi\_lens : Provides a differentiable log normal mass map simulator specifically designed for for weak lensing inference. Then you can describe a bit the simulations like tomographic bins etc I guess

    - We construct our ground truth

    - We use implicit inference techniques to go from the compressed stat to the posteriors}
    }
  % results heading (mandatory)
   {Our experiments validate the effectiveness of LFI methods,  as we demonstrate comparable posterior distributions between LFI and Bayesian forward modeling approaches \justine{field-based Bayesian Hierarchical Model analysis}. Additionally, our results provide insights into the design of data compression procedures to derive low-dimensional summary statistics while minimizing information loss. All codes associated with this paper can be found at this link \github.}
  % conclusions heading (optional), leave it empty if necessary 
   {}
   \keywords{methods: statistical – gravitational lensing: weak – cosmology: large-scale structure of Universe
               }

@@

In this paper, we present a comparison of the performance of various neural network-based compression schemes within the Likelihood-Free Inference (LFI) framework. These methods differ in terms of the loss functions used to train the neural network, but they share the same inference strategy based on neural density estimation. To address the need for result validation in the LFI context, we employ a Bayesian forward modeling approach that enables inference on the joint posterior of the cosmological parameters and the convergence field. The forward model is based on \href{https://github.com/DifferentiableUniverseInitiative/sbi_lens}{\url{SbiLens}}, a Python package designed for weak lensing implicit inference with differentiable simulator implemented in Jax. \texttt{SbiLens} allows the sampling of convergence maps in a tomographic setting, accounting for the cross-correlation between different redshift bins.

The paper is structured as follows: in \autoref{Sec:Motivation}, we illustrate the motivation behind this work. In \autoref{Sec:the SBILens framework}, we introduce the \texttt{SbiLens} framework and describe the simulated data used in this work. In \autoref{Sec:experiment}, we detail the inference strategy and the three different approaches we used: the power spectrum, the map-based inference based on the Bayesian forward-modeling, and the map-based inference based on the LFI. In the same section, we also provide a detailed overview of different neural compression strategies. In \autoref{Sec:results}, we discuss the results and validate the LFI approaches. Finally, we conclude in \autoref{Sec:conclusion}. 
%------------------------------------------------------------------
%------------------------------------------------------------------
%------------------------------------------------------------------
%------------------------------------------------------------------
\section{Motivation}\label{Sec:Motivation}
%--------------------------------------------------------------------
With the increased statistical power of stage IV surveys, our cosmological analysis should not rely on the measurement of sub-optimal summary statistics, that may not fully capture the non-Gaussian information present in the lensing field at the scales accessible to future surveys. In this paper we introduce a forward model that directly extracts information from the raw pixel data, rather than relying on the analytical evaluation of summary statistics. By doing so, we aim to preserve all available information and facilitate the incorporation of systematic effects and the combination of multiple cosmological probes through joint simulations.
In this context, the simulator of the observables serves as our physical model, where each component is tractable. These models, often refereed to as \textit{probabilistic program}, can be described as follows. They take as input a vector parameter $\bm{\theta}$, which describes the underlying deterministic model. Then, they sample internal states, dubbed \textit{latent variables}, from the distribution $\bm{z}_i \sim p_i(\bm{z}_i|\bm \theta, \bm{z}_{<i})$, than can be directly or indirectly related to a physically meaningful state of the system. Finally, they generate the output $\bm x$ from the distribution $\ p(\bm d|\bm \theta)$, where $\bm{d}$ represents the observations.

The ultimate goal of Bayesian inference in cosmology is to compute the posterior distribution:
\begin{equation}\label{Eq:posterior}
     p(\bm{\theta}|\bm{x})= 
     \frac{p(\bm{x}|\bm{\theta})p(\bm{\theta})}
     {\int d\bm{\theta'}p(\bm{x}|\bm{\theta}')p(\bm{\theta}')},
\end{equation}
however, a problems arises because the marginal likelihood $p(\bm{x}|\bm{\theta})$ is typically intractable:
\begin{equation}
    p(\bm{x}|\bm{\theta})=\int p(\bm{x},\bm{z}|\bm{\theta}) d\bm{z}=\int p(\bm{x}|\bm{z},\bm{\theta})p(\bm{z}|\bm{\theta}) d\bm{z}.
\end{equation}
as it corresponds to an integral over all possible trajectories throughout the latent space.
To overcome this limitation while still capturing the full information content of the data, two different approaches have been proposed in the literature. Although these approaches are often referred to by different names, in this work, we will make the following distinction:
%--------------------------------------------------------------------
\paragraph{\textbf{Explicit inference}} referring to all Likelihood-based inference approaches.  
This approach involves treating the simulator as a probabilistic model and performing inference over the joint posterior:
\begin{equation}
        p(\bm{\theta},\bm{z}|\bm{x})\propto  p(\bm{x},\bm{z}|\bm{\theta}) p(\bm{z}|\bm{\theta}).
\end{equation}
In the context of the forward model and full field analysis, fall in this category the Bayesian Hierarchical Models (BHMs).
A Bayesian forward approach involves using a given physical model to predict observations and then comparing these predictions with observations to infer the parameters of the model.
However, performing a global analysis on complex and large datasets, such as a lensing dataset, is not feasible. Therefore, the idea of dividing the global analysis into several subgroups, analyzing them in multiple steps, and using the output of each of those as input for the next one. From here, the \textit{hierarchical} nature of this methodology.
%--------------------------------------------------------------------
\paragraph{\textbf{Implicit inference}} referring to all the approaches not relying on an analytical model to describe the signal, but rather on learning the likelihood from simulations. This second class of approaches involves treating the simulator as a black box with only the ability to sample from the joint distribution:
\begin{equation}
    (\bm{x}, \bm{\theta})\sim p(\bm{x}, \bm{\theta}).
\end{equation}
Within this class of methods, we can differentiate between more traditional methods such as Approximate Bayesian Computation (ABC) and Density Estimation Likelihood-Free Inference (DELFI) methods. ABC employs rejection-criteria based approaches to approximate the likelihood by comparing simulations with data. 
In this work, our focus will be on DELFI methods, which approach the inference task as a density estimation problem. \\
We can divide Likelihood-Free Inference approaches into two separate steps:
\begin{enumerate}
    \item  Automatically learning an optimal low-dimensional summary statistic.
    \item Using Neural Density Estimation in low dimensions to infer the target distributions.
\end{enumerate}
In the first step, we introduce a parametric function $f_{\varphi}$ such that:
     \begin{equation}
         \bm{y}=f_{\varphi}(\bm{\theta}),
     \end{equation}
which aims to reduce the dimensionality of the data while preserving information. Typically, the compressed statistics $\bm{y}$ is assumed to have the same dimension of $\bm{\theta}$. \\
In the second step, Neural Density Estimation can target either building an estimate $p_{\phi}$ of the likelihood function $p(\bm{x}|\bm{\theta})$ (referred to as the Neural Likelihood Estimation (NLE) task), or targeting the posterior distribution $p(\bm{\theta}|\bm{x})$, (known as Neural Posterior Estimation (NPE) task).

It is important to note that not all neural compression techniques are equivalent. Many papers that apply neural techniques for extracting summary statistics have used sub-optimal compression techniques, such as Mean Square Error or Maximum Absolute Error.
However, it is also important to consider that in this case, the compression of summaries will be sub-optimal, resulting in a loss of information (larger contours), but it will not introduce bias in the final constraints. On the contrary, many other papers rely on assuming proxy Gaussian likelihoods and estimate the mean and covariance of these likelihoods from simulations. While likelihood-free inference using MSE and MAE summaries is robust in the sense that the approximations made in the compression steps will not introduce bias to the parameter inferences, the same cannot be said for summaries obtained from Gaussian likelihood approximations.
We summarize the different neural compression strategies found in the literature in \autoref{tab:biblio_survey}.

\vspace{1cm}
Ultimately, for a given simulation model, if an optimal compression statistic is used, the two approaches should converge to the same posterior. Therefore, the goals of this paper will be:
\begin{enumerate}
    \item Find the optimal compression strategy.
    \item Demonstrate that by using this optimal compression strategy, both the implicit and explicit methods yield comparable results.
\end{enumerate}

%------------------------------------------------------------------
%------------------------------------------------------------------
%------------------------------------------------------------------
%------------------------------------------------------------------
\section{The SBILens framework}\label{Sec:the SBILens framework}
\denise{PEOPLE CAN READ THIS!!}
\subsection{Lognormal Modeling}\label{Sec:Lognormal Modeling}
For various cosmological applications, the non-Gaussian field can be modeled as a Lognormal field \citep{coles1991lognormal,bohm2017bayesian}.
This model offers the advantage of generating the matter or convergence field rapidly while allowing the extraction of information beyond the two-point statistics. 
Although studies demonstrated that this model fails in describing the 3D field \citep{klypin2018density}, it properly describes the 2D convergence field \citep{clerkin2017testing, xavier2016improving}.
Assuming a simulated Gaussian convergence map $\kappa_g$, whose statistical properties are fully described by its power spectrum $C_{\ell}$ we know that this model is not a suitable representation of late-time and more evolved structures. One potential solution is to find a transformation $f(\kappa_g)$ of this map that captures the non-Gaussian features in the convergence field. In doing so, it is crucial to ensure that the transformed map maintains the correct mean and variance, effectively recovering the correct two-point statistics.
Denoting $\mu$ and $\sigma_g^2$ the mean and covariance matrix of $\kappa_g$ respectively, we can define the transformed convergence $\kappa_{ln}$ as a shifted lognormal random field:
\begin{equation}\label{Eq:log_norm_kappa}
    \kappa_{ln}=e^{\kappa_{g}}-\lambda, 
\end{equation}
where $\lambda$ is a free parameter that determines the shift of the lognormal distribution. The convergence $\kappa$ in a given redshift bin is fully determined by the shift parameter $\lambda$, the mean $\mu$ of the associated Gaussian field $\kappa_g$, and its variance $\sigma_{g}^2\equiv \xi_g$.
The correlation of the lognormal field, denoted as $\xi_{ln}$, is also a function of these variables and is related to $\xi^{ij}_g$ through the following equations:
\begin{align}
    \xi^{ij}_{ln}(\theta) & \equiv \lambda_i \lambda_j (e^{ \xi^{ij}_g(\theta)}-1) \nonumber \\ 
    \xi^{ij}_g(\theta)&=\log{\left[ \frac{\xi^{ij}_{ln}(\theta)}{\lambda_i \lambda_j}+1\right ]}. \label{Eq:log_norm_corr}
\end{align}
Here $i$ and $j$ define a pair of redshift bins.
The parameter $\lambda$, also known as \textit{minimum convergence
parameter}, defines the lowest values for all possible values of $\kappa$.
The modeling of the shift parameter can be approached in various ways. For example, it can be determined by matching moments of the distribution \citep{xavier2016improving} or by treating it as a free parameter \citep{hilbert2011cosmic}. In general, the value of $\lambda$ depends on the redshift, cosmology, and the scale of the field at which smoothing is applied.

@@

 In the SbiLens framework, the sampling of the convergence maps can be described as follows. 

@@

     \bm{\kappa_g}=\bm{Z}*\tilde{\bm{\Sigma} }

@@

where $\bm{Z}$ represents the latent variables of the simulator.

@@

@@

To simulate our data, we develop the SbiLens package, which employs a Lognormal model to represent the convergence maps, as explained in the previous section. Specifically, the package uses the public library \href{https://github.com/DifferentiableUniverseInitiative/jax_cosmo}{\texttt{jax-cosmo}} \citep{Campagne_2023} to compute the theoretical power- and cross-spectra. The computation of the lognormal shift parameter is performed using the \texttt{Cosmomentum} code \citep{friedrich2018density, friedrich2020primordial}, which utilizes perturbation theory to compute the cosmology-dependent shift parameters. In Cosmomentum the calculation of the shift parameters assumes a cylindrical window function, while our pixels are rectangular. Following \citet{boruah2022map}, we compute the shift parameters at a characteristic scale, $R=\Delta L/\pi$, where $\Delta L$ represents the pixel resolution. The shift parameters at each redshift are computed using the fiducial cosmology values for $\Omega_b$, $h_0$, and $n_s$. To incorporate the cosmology dependence of $\lambda$ from $\Omega_c$, $\sigma_8$ and $w_0$, we calculate the shift for different points in the cosmological parameter space and then interpolate the shift values for other points in the parameter space.

As mentioned by \citet{boruah2022map}, the perturbation theory-based approach may not provide accurate results at small scales. However, the main objective of this paper is to compare different inference strategies and we are not concerned with the potential implications of this approximation. 
In future applications of the presented inference pipeline, the simulation procedure will be based on N-body simulations.

Each map is reproduced on a regular grid with dimensions of $256 \times 256$ pixels and covers an area of $10\times 10$ deg$^2$.

@@

where we set the shape noise $\sigma_e = 0.26$ and the galaxy number density $n_{gal}=27$ arcmin$^{-2}$. Both the shape noise and galaxy number density values are obtained from SRD. The pixel area is given by $A_{pix}\approx$. 

@@

@@

@@

To obtain the probability distribution of the cosmological parameters using a summary-statistics-methods, we focus on the 2-point statistics, specifically, on the angular power spectra $C_{\ell}$.
    \log{\mathcal{L}}(\bm{\theta})=-\frac{1}{2}[\bm{d}-\bm{\mu}(\bm{\theta})]^{T}\bm{C}^{-1}[\bm{d}-\bm{\mu}(\bm{\theta})].
To compute the expected theoretical predictions $\bm{\mu}(\bm{\theta})$ we use the public library \href{https://github.com/DifferentiableUniverseInitiative/jax_cosmo}{\texttt{jax-cosmo}}. 
    \text{Cov}(C_{\ell},C_{\ell}')=\frac{1}{f_{sky}(2 \ell+1)}\left(C_{\ell}+\frac{\sigma_{\epsilon}^2}{2n_s}\right)

@@

As mentioned before, this means that the simulator serves as the physical deterministic model capable of generating the non-linear representation of the convergence map.

However, in practical terms, the measurement of convergence for each pixel and bins will differ from real observations due to noise. This is taken into consideration in the likelihood. Specifically, for LSST Y10, the number of galaxies for each pixel should be sufficiently high so that, according to the central limit theorem, we can assume the observation is characterized by Gaussian noise, with $\sigma_n^2=\sigma_e^2/N_s$, where $N_s$ represents the total number of source galaxies per bin and pixel. Given $\sigma_n^2$ the variance of this Gaussian likelihood, its log-form can be expressed as:

@@

    =\sum_i^{N_{pix}} \sum_{j}^{N_{bins}}\frac{[\kappa_{i,j}-\kappa^{obs}_{i,j}]^2}{2\sigma_n^2},

@@

Since the full-field approach does not rely on any summary statistics, it typically leads to a high-dimensional problem, requiring more sophisticated statistical techniques. To sample the posterior distribution for $\bm{\theta}$, we use a Hamiltonian Monte Carlo (HMC) algorithm. The HMC algorithm is particularly helpful in high-dimensional spaces where a large number of steps are required to effectively explore the space. It improves the sampling process by leveraging the information contained in the gradients to guide the sampling process. As the code is implemented in Jax, the gradients are accessible via automatic differentiation. 

>>>>

  \abstract
  % context heading (optional)
  % {} leave it empty if necessary  
   {In many cosmological applications, the likelihood function is either unknown or intractable, limiting the use of traditional inference methods. Likelihood-free inference (LFI) approaches offer a promising solution by using forward-simulated synthetic data to approximate the posterior (or likelihood) distribution instead of evaluating an explicit likelihood function. However, these methods suffer from the curse of dimensionality, necessitating the development of compression schemes to reduce high-dimensional data into lower-dimensional summary statistics.
   }   
  % aims heading (mandatory)
    {The goal of this paper is to investigate the performance of several neural compression strategies for full field Likelihood free inference by density estimation. Moreover, we aim to demonstrate that, by using an optimal compression strategy, the posterior distribution is comparable to those derived from a Bayesian forward modeling approach.}  
  % methods heading (mandatory)
    { We present a comparative analysis of different loss functions employed during the training of the neural network. We evaluate their performance by measuring their impact on the constraints on the $\Lambda$CDM parameters expected from LSST-Y10.
    For both forward-modeling the convergence field and generating observed mock data, we develop the \href{https://github.com/DifferentiableUniverseInitiative/sbi_lens}{\url{SBILens}} package. 
    \texttt{SBILens} is a Jax-based differentiable physical model used to generate convergence maps as lognormal random fields. It is specifically designed for likelihood-free inference and Bayesian inference algorithms that require access to likelihood derivatives.}

   {Our experiments validate the effectiveness of LFI methods, as we demonstrate comparable posterior distributions between LFI and Bayesian forward modeling approaches.
   All codes associated with this paper can be found at this link \github.}
   {}
   \keywords{methods: statistical – gravitational lensing: weak – cosmology: large-scale structure of Universe
               }

@@

In this paper, we present a comparison of the performance of various neural network-based compression schemes within the Likelihood-Free Inference (LFI) framework. These methods differ in terms of the loss functions used to train the neural network, but they share the same inference strategy based on neural density estimation. To address the need for result validation in the LFI context, we employ a Bayesian forward modeling approach that enables inference on the joint posterior of the cosmological parameters and the convergence field. The forward model is based on \href{https://github.com/DifferentiableUniverseInitiative/sbi_lens}{\url{SBILens}}, a Python package designed for weak lensing inference with differentiable simulator implemented in Jax. \texttt{SBILens} allows the sampling of convergence maps in a tomographic setting, accounting for the cross-correlation between different redshift bins.

The paper is structured as follows: in \autoref{Sec:Motivation}, we illustrate the motivation behind this work. In \autoref{Sec:the SBILens framework}, we introduce the \texttt{SBILens} framework and describe the simulated data used in this work. In \autoref{Sec:experiment}, we detail the inference strategy and the three different approaches we used: the power spectrum, the map-based inference based on the Bayesian forward-modeling, and the map-based inference based on the LFI. In the same section, we also provide a detailed overview of different neural compression strategies. In \autoref{Sec:results}, we discuss the results and validate the LFI approaches. Finally, we conclude in \autoref{Sec:conclusion}. 
%------------------------------------------------------------------
%------------------------------------------------------------------
%------------------------------------------------------------------
%------------------------------------------------------------------
\section{Motivation}\label{Sec:Motivation}
%--------------------------------------------------------------------
With the increased statistical power of stage IV surveys, our cosmological analysis should not rely on the measurement of sub-optimal summary statistics, which may not fully capture the non-Gaussian information present in the lensing field at the scales accessible to future surveys. In this paper, we introduce a forward model that directly extracts information from the convergence maps, rather than relying on the analytical evaluation of summary statistics. By doing so, we aim to preserve all available information and facilitate the incorporation of systematic effects and the combination of multiple cosmological probes through joint simulations.
In a forward model context, the simulator of the observables serves as our physical model, where each component is tractable. These models, often referred to as \textit{probabilistic program}, as illustrated by \citet{cranmer2020frontier}, can be described as follows: The models take as input a vector parameter $\bm{\theta}$, which describes the underlying deterministic model. Then, they sample internal states $\bm{z}$, dubbed \textit{latent variables}, from the distribution $p(\bm{z}|\bm \theta)$. These states can be directly or indirectly related to a physically meaningful state of the system. Finally, the models generate the output $\bm x$ from the distribution $p(\bm x|\bm \theta,\bm{z})$, where $\bm{x}$ represents the observations.

The ultimate goal of Bayesian inference in cosmology is to compute the posterior distribution:
\begin{equation}\label{Eq:posterior}
     p(\bm{\theta}|\bm{x})= 
     \frac{p(\bm{x}|\bm{\theta})p(\bm{\theta})}
     {\int d\bm{\theta'}p(\bm{x}|\bm{\theta}')p(\bm{\theta}')},
\end{equation}
however, a problems arises because the marginal likelihood $p(\bm{x}|\bm{\theta})$ is typically intractable:
\begin{equation}
    p(\bm{x}|\bm{\theta})=\int p(\bm{x},\bm{z}|\bm{\theta}) d\bm{z}=\int p(\bm{x}|\bm{z},\bm{\theta})p(\bm{z}|\bm{\theta}) d\bm{z}.
\end{equation}
since it involves integrating over all potential paths through the latent space.
To overcome this limitation while still capturing the full information content of the data, two different approaches have been proposed in the literature. Although these approaches are often referred to by different names, hereinafter we will make the following distinction:
%--------------------------------------------------------------------
\paragraph{\textbf{Explicit inference}} referring to all Likelihood-based inference approaches.  
This approach involves treating the simulator as a probabilistic model and performing inference over the joint posterior:
\begin{equation}
        p(\bm{\theta},\bm{z}|\bm{x})\propto  p(\bm{x},\bm{z}|\bm{\theta}) p(\bm{z}|\bm{\theta}).
\end{equation}
In the context of the forward model and full-field analysis, Bayesian Hierarchical Models (BHMs) fall into this category.
A Bayesian forward approach involves using a given physical model to predict observations and then comparing these predictions with observations to infer the parameters of the model.
However, performing a global analysis on complex and large datasets, such as a lensing dataset, is not feasible. Therefore, the idea of dividing the global analysis into several subgroups, analyzing them in multiple steps, and using the output of each of those as input for the next one. From here, the \textit{hierarchical} nature of this methodology.
%--------------------------------------------------------------------
\paragraph{\textbf{Implicit inference}} referring to all the approaches that try to infer the marginal likelihood from simulations. This second class of approaches involves treating the simulator as a black box with only the ability to sample from the joint distribution:
\begin{equation}
    (\bm{x}, \bm{\theta})\sim p(\bm{x}, \bm{\theta}).
\end{equation}
Within this class of methods, we can differentiate between more traditional methods such as Approximate Bayesian Computation (ABC) and Density Estimation Likelihood-Free Inference (DELFI) methods. ABC employs rejection-criteria based approaches to approximate the likelihood by comparing simulations with data. 
In this work, our focus will be on DELFI methods, which approach the inference task as a density estimation problem. \\
In particular, the standard deep learning approach for Likelihood-Free Inference can be described as separated into two distinct steps:
\begin{enumerate}
    \item  Automatically learning an optimal low-dimensional summary statistic.
    \item Using Neural Density Estimation in low dimensions to infer the target distributions.
\end{enumerate}
In the first step, we introduce a parametric function $F_{\varphi}$ such that:
     \begin{equation}
         \bm{t}=F_{\varphi}(\bm{\theta}),
     \end{equation}
which aims to reduce the dimensionality of the data while preserving information. Typically, the compressed statistics $\bm{t}$ is assumed to have the same dimension of $\bm{\theta}$. \\
In the second step, Neural Density Estimation can target either building an estimate $p_{\varphi}$ of the likelihood function $p(\bm{x}|\bm{\theta})$ (referred to as the Neural Likelihood Estimation (NLE) task), or targeting the posterior distribution $p(\bm{\theta}|\bm{x})$, (known as Neural Posterior Estimation (NPE) task).

The main motivation for this work is to evaluate the impact of a given compression strategy on the final posterior distribution and determine whether an optimal compression strategy exists. Indeed, it is important to consider that different neural compression techniques may not be equivalent. We summarize the various neural compression strategies found in the literature in \autoref{tab:biblio_survey}. 
Many of these papers have used neural compression techniques that rely on optimizing the Mean Square Error or the Maximum Absolute Error. \\
As we will demonstrate in the following sections, this corresponds to training the model to estimate the mean and the median of the posterior distribution. Other papers rely on assuming proxy Gaussian likelihoods and estimate the mean and covariance of these likelihoods from simulations.
Such compression of summaries could be sub-optimal in certain applications, resulting in a loss of information and hence larger contours.

Therefore, our primary focus will be to investigate the effectiveness and validity of these compression techniques in the context of weak lensing applications (within the limits of our setting).

\vspace{1cm}
Ultimately, we should keep in mind that, for a given simulation model, if an optimal compression statistic is used, the two approaches should converge to the same posterior. Therefore, the goals of this work will be:
\begin{enumerate}
    \item Find a near-optimal compression strategy.
    \item Demonstrate that by using this near-optimal compression strategy, both the implicit and explicit methods yield comparable results.
\end{enumerate}

%------------------------------------------------------------------
%------------------------------------------------------------------
%------------------------------------------------------------------
%------------------------------------------------------------------
\section{The SBILens framework}\label{Sec:the SBILens framework}
To investigate the questions above, we have developed the Python package \href{https://github.com/DifferentiableUniverseInitiative/sbi_lens}{\texttt{SBILens}}, a weak lensing differentiable simulator based on a log-normal model. \texttt{SBILens} enables the sampling of convergence maps in a tomographic setting while considering the cross-correlation between different redshift bins.
\subsection{Lognormal Modeling}\label{Sec:Lognormal Modeling}
For various cosmological applications, the non-Gaussian field can be modeled as a Lognormal field \citep{coles1991lognormal,bohm2017bayesian}.
This model offers the advantage of generating the matter or convergence field rapidly while allowing the extraction of information beyond the two-point statistics. 
Although studies demonstrated that this model fails in describing the 3D field \citep{klypin2018density}, it properly describes the 2D convergence field \citep{clerkin2017testing, xavier2016improving}.
Assuming a simulated Gaussian convergence map $\kappa_g$, whose statistical properties are fully described by its power spectrum $C_{\ell}$ we know that this model is not a suitable representation of late-time and more evolved structures. One potential solution is to find a transformation $f(\kappa_g)$ of this map that captures the non-Gaussian features in the convergence field. In doing so, it is crucial to ensure that the transformed map maintains the correct mean and variance, effectively recovering the correct two-point statistics.
Denoting $\mu$ and $\sigma_g^2$ the mean and covariance matrix of $\kappa_g$ respectively, we can define the transformed convergence $\kappa_{ln}$ as a shifted lognormal random field:
\begin{equation}\label{Eq:log_norm_kappa}
    \kappa_{ln}=e^{\kappa_{g}}-\lambda, 
\end{equation}
where $\lambda$ is a free parameter that determines the shift of the lognormal distribution. The convergence $\kappa$ in a given redshift bin is fully determined by the shift parameter $\lambda$, the mean $\mu$ of the associated Gaussian field $\kappa_g$, and its variance $\sigma_{g}^2$.
The correlation of the lognormal field, denoted as $\xi_{ln}$, is also a function of these variables and is related to $\xi^{ij}_g$ through the following equations:
\begin{align}
    \xi^{ij}_{ln}(\theta) & \equiv \lambda_i \lambda_j (e^{ \xi^{ij}_g(\theta)}-1) \nonumber \\ 
    \xi^{ij}_g(\theta)&=\log{\left[ \frac{\xi^{ij}_{ln}(\theta)}{\lambda_i \lambda_j}+1\right ]}. \label{Eq:log_norm_corr}
\end{align}
Here $i$ and $j$ define a pair of redshift bins.
The parameter $\lambda$, also known as \textit{minimum convergence
parameter}, defines the lowest values for all possible values of $\kappa$.
The modeling of the shift parameter can be approached in various ways. For example, it can be determined by matching moments of the distribution \citep{xavier2016improving} or by treating it as a free parameter \citep{hilbert2011cosmic}. In general, the value of $\lambda$ depends on the redshift, cosmology, and the scale of the field at which smoothing is applied.

@@

 In the SBILens framework, the sampling of the convergence maps can be described as follows. 

@@

     \bm{\kappa_g}=\hat{\bm{Z}}\cdot\tilde{\bm{\Sigma} }

@@

where $\hat{\bm{Z}}$ represents the Fourier transform of the latent variables of the simulator.

@@

To ensure that we recover the correct auto- and cross-power spectra, we compare the results from our simulations to theoretical predictions for different tomographic bin combinations. We show the results in \autoref{fig:psconvergence_maps}. \denise{Not sure we want this}

@@

To simulate our data, we develop the SBILens package, which employs a Lognormal model to represent the convergence maps, as explained in the previous section. Specifically, the package uses the public library \href{https://github.com/DifferentiableUniverseInitiative/jax_cosmo}{\texttt{jax-cosmo}} \citep{Campagne_2023} to compute the theoretical power- and cross-spectra. The computation of the lognormal shift parameter is performed using the \texttt{Cosmomentum} code \citep{friedrich2018density, friedrich2020primordial}, which utilizes perturbation theory to compute the cosmology-dependent shift parameters. In Cosmomentum the calculation of the shift parameters assumes a cylindrical window function, while our pixels are rectangular. Following \citet{boruah2022map}, we compute the shift parameters at a characteristic scale, $R=\Delta L/\pi$, where $\Delta L$ represents the pixel resolution. \\
For each redshift bin, we tested the dependency of the shift parameter $\lambda$ on various cosmological parameters. Specifically, we investigated how the value of $\lambda$ changed when varying a specific cosmological parameter while keeping the others fixed.
Our findings revealed that the parameters $\Omega_b$, $h_0$, and $n_s$ had almost no significant impact on $\lambda$. As a result, we computed the shift parameters for each redshift using the fiducial cosmology values of $\Omega_b$, $h_0$, and $n_s$.
To account for the cosmology dependence of $\lambda$ on $\Omega_c$, $\sigma_8$, and $w_0$, we calculated the shift for various points in the cosmological parameter space and then interpolated the shift values for other points in the parameter space.
Each map is reproduced on a regular grid with dimensions of $256 \times 256$ pixels and covers an area of $10\times 10$ deg$^2$.  An example of a tomographic convergence map simulated using the \texttt{SBILens} package is shown in \autoref{fig:convergence_maps}. \denise{TODO: change disp images}
% ############# PRIOR COSMO TABLE  #############

@@

where we set the shape noise $\sigma_e = 0.26$ and the galaxy number density $n_{gal}=27$ arcmin$^{-2}$. Both the shape noise and galaxy number density values are obtained from SRD. The pixel area is given by $A_{pix}\approx 5.49$ arcmin$^2$. 

@@

% ############# PLOT CONVERGENCE MAPS  #############
    \includegraphics[width=\textwidth]{figures/bin1and2.pdf}
    \includegraphics[width=\textwidth]{figures/bin3.pdf}
    \includegraphics[width=0.5\textwidth]{figures/bin5.pdf}
     Example of convergence map simulated using the \texttt{SBILens} package.
     \label{fig:convergence_maps}
% ############# PLOT CONVERGENCE MAPS  #############
    \includegraphics[width=\textwidth]{figures/psconvergencemaps.pdf}
    Convergence power spectra for different tomographic bin combinations. The solid yellow line shows the measurement from 20 simulated maps using the survey setting described in \autoref{Sec:the SBILens framework}, while the black dashed line shows the theoretical predictions computed using jax-cosmo. In this figure, the shaded regions represent the standard deviation from 20 independent map realizations.
     \label{fig:psconvergence_maps}

@@

In the following section, we illustrate the inference strategy for the two approaches under investigation: the Bayesian forward modeling and the map-based inference based on the LFI.
Additionally, we conduct a power spectrum study. Indeed, as discussed in \autoref{Sec:Lognormal Modeling}, log-normal fields offer the advantage of rapidly generating convergent fields while accounting for non-Gaussianities. To emphasize this claim further, along with the full-field analysis, we include a power spectrum analysis. This analysis demonstrates that there is indeed a gain of information when adopting a full-field approach. \\
Finally, we provide a detailed overview of various neural compression strategies that differ in terms of the loss functions used to train the neural network.

@@

@@

As mentioned before, this means that the simulator serves as the physical model capable of generating the non-linear representation of the convergence map.

However, in practical terms, the measurement of convergence for each pixel and bin will differ from real observations due to noise. This is taken into consideration in the likelihood. Specifically, for LSST Y10, the number of galaxies for each pixel should be sufficiently high so that, according to the central limit theorem, we can assume the observation is characterized by Gaussian noise, with $\sigma_n^2=\sigma_e^2/N_s$, where $N_s$ represents the total number of source galaxies per bin and pixel. Given $\sigma_n^2$ the variance of this Gaussian likelihood, its negative log-form can be expressed as:

@@

    =-\sum_i^{N_{pix}} \sum_{j}^{N_{bins}}\frac{[\kappa_{i,j}-\kappa^{obs}_{i,j}]^2}{2\sigma_n^2},

@@

Since the full-field approach does not rely on any summary statistics, it typically leads to a high-dimensional problem, requiring more sophisticated statistical techniques. To sample the posterior distribution for $\bm{\theta}$, we use a Hamiltonian Monte Carlo (HMC) algorithm. 
Specifically, we employ the \texttt{NUTS} algorithm, an adaptive variant of HMC implemented in \href{https://github.com/pyro-ppl/numpyro}{\texttt{NumPyro}} \citep{phan2019composable, bingham2019pyro}, which uses the No U-Turn Sampler (NUTS, \citet{hoffman2014no}). \\
The HMC algorithm is particularly helpful in high-dimensional spaces where a large number of steps are required to effectively explore the space. It improves the sampling process by leveraging the information contained in the gradients to guide the sampling process. As the code is implemented in Jax, the gradients are accessible via automatic differentiation. 
To obtain the probability distribution of the cosmological parameters using a summary-statistics method, we focus on the 2-point statistics, specifically, on the angular power spectra $C_{\ell}$.
    \mathcal{L}(\bm{\theta})=-\frac{1}{2}[\bm{d}-\bm{\mu}(\bm{\theta})]^{T}\bm{C}^{-1}[\bm{d}-\bm{\mu}(\bm{\theta})].
To compute the expected theoretical predictions $\bm{\mu}(\bm{\theta})$ we use the public library \href{https://github.com/DifferentiableUniverseInitiative/jax_cosmo}{\texttt{jax-cosmo}} \citep{Campagne_2023}. 
